---
title: "Higher Education Students Performance Evaluation"
output: 
  html_document:
    theme: darkly
    df_print: paged
    highlight: breezedark
    css: styles.css
    fig_width: 7
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE) 
```
This is a classification analysis to try determining the grade of a student based on other attributes.

Dataset source: https://www.kaggle.com/csafrit2/higher-education-students-performance-evaluation

## Loading Libraries

```{r}
library(dplyr) # data preprocessing
library(ggplot2) # data visualization
library(e1071) # svm
library(randomForest) # random forest
library(nnet) # multinomial logistic regression
library(glmnet) # the lasso
library(class) # KNN
library(caret) # CV
library(ggdark)
```

## Reading the data and viewing the first few rows

```{r}
data <- read.csv("./data/student_prediction.csv")
head(data)
```


<br/>
Many of the columns are using number to indicate factor type of data, therefore, each column will be
updated to for better representation and understanding for me.
<br/>

```{r}
data <- data %>%
  rename(STUDENT_ID = `ï..STUDENTID`) %>%
  mutate(AGE = recode(AGE, `1` = "18 - 21", `2` = "22 - 25", `3` = "> 26")) %>%
  mutate(GENDER = recode(GENDER, `1` = "f", `2` = "m")) %>%
  mutate(HS_TYPE = recode(HS_TYPE, `1` = "private", `2` = "state", `3` = "other")) %>%
  mutate(SCHOLARSHIP = recode(SCHOLARSHIP, `1` = 0, `2` = 25, `3` = 50, `4` = 75, `5` = 100)) %>%
  mutate(WORK = recode(WORK, `1` = TRUE, `2` = FALSE)) %>%
  mutate(ACTIVITY = recode(ACTIVITY, `1` = TRUE, `2` = FALSE)) %>%
  mutate(PARTNER = recode(PARTNER, `1` = TRUE, `2` = FALSE)) %>%
  mutate(SALARY = recode(SALARY, `1` = "135 - 200", `2` = "201 - 270", 
                         `3` = "271 - 340", `4` = "341 - 410", `5` = "> 410")) %>%
  mutate(TRANSPORT = recode(TRANSPORT, `1` = "Bus", `2` = "Private car / taxi",
                            `3` = "Bicycle", `4` = "Other")) %>%
  mutate(LIVING = recode(LIVING, `1` = "rental", `2` = "dormitory", 
                         `3` = "with family", `4` = "other")) %>%
  mutate(MOTHER_EDU = recode(MOTHER_EDU, `1` = "primary school", `2` = "secondary school",
                             `3` = "high school", `4` = "university", 
                             `5` = "MSc", `6` = "PhD")) %>%
  mutate(FATHER_EDU = recode(FATHER_EDU, `1` = "primary school", `2` = "secondary school",
                             `3` = "high school", `4` = "university", 
                             `5` = "MSc", `6` = "PhD")) %>%
  rename(PARENTAL_STATUS = KIDS) %>%
  mutate(PARENTAL_STATUS = recode(PARENTAL_STATUS, `1` = "married", 
                                  `2` = "divorced", `3` = "at least one died")) %>%
  mutate(MOTHER_JOB = recode(MOTHER_JOB, `1` = "retired", `2` = "housewife",
                             `3` = "government officer", `4` = "private sector employee", 
                             `5` = "self-employment", `6` = "other")) %>%
  mutate(FATHER_JOB = recode(FATHER_JOB, `1` = "retired",
                             `2` = "government officer", `3` = "private sector employee", 
                             `4` = "self-employment", `5` = "other")) %>%
  mutate(STUDY_HRS = recode(STUDY_HRS, `1` = "None", `2` = "<5", `3` = "6 - 10",
                            `4` = "11 - 20", `5` = "> 20")) %>%
  mutate(READ_FREQ = recode(READ_FREQ, `1` = "None", `2` = "Sometimes", `3` = "Often")) %>%
  mutate(READ_FREQ_SCI = recode(READ_FREQ_SCI, `1` = "None", `2` = "Sometimes", `3` = "Often")) %>%
  mutate(ATTEND_DEPT = recode(ATTEND_DEPT, `1` = TRUE, `2` = FALSE)) %>%
  rename(ATTEND_SEMINAR = ATTEND_DEPT) %>%
  mutate(IMPACT = recode(IMPACT, `1` = "Positive", `2` = "Negative", `3` = "Neutral")) %>%
  mutate(ATTEND = recode(ATTEND, `1` = "always", `2` = "sometimes", `3` = "never")) %>%
  mutate(PREP_STUDY = recode(PREP_STUDY, `1` = "alone", `2` = "with friends", `3` = "not applicable")) %>%
  mutate(PREP_EXAM = recode(PREP_EXAM, `1` = "close to exam", `2` = "regularly", `3` = "never")) %>%
  mutate(NOTES = recode(NOTES, `1` = "never", `2` = "sometimes", `3` = "always")) %>%
  mutate(LISTENS = recode(LISTENS, `1` = "never", `2` = "sometimes", `3` = "always")) %>%
  mutate(LIKES_DISCUSS = recode(LIKES_DISCUSS, `1` = "never", `2` = "sometimes", `3` = "always")) %>%
  mutate(CLASSROOM = recode(CLASSROOM, `1` = "not useful", `2` = "useful", `3` = "not applicable")) %>%
  mutate(CUML_GPA = recode(CUML_GPA, `1` = "< 2.00", `2` = "2.00 - 2.49", `3` = "2.50 - 2.99",
                           `4` = "3.00 - 3.49", `5` = "> 3.49")) %>%
  mutate(EXP_GPA = recode(EXP_GPA, `1` = "< 2.00", `2` = "2.00 - 2.49", `3` = "2.50 - 2.99",
                           `4` = "3.00 - 3.49", `5` = "> 3.49")) %>%
  mutate(GRADE = recode(GRADE, `0` = "fail", `1` = "DD", `2` = "DC", `3` = "CC", `4` = "CB",
                        `5` = "BB", `6` = "BA", `7` = "AA")) %>%
  select(-STUDENT_ID) # removing unique value column
  
head(data)
```

<br/>
much better, now change most of the columns to factor
<br/>

```{r}
col_names <- names(data)
factor_columns <- col_names[!col_names %in% c("STUDENT_ID", "WORK", "ACTIVITY", "PARTNER",
                                              "X._SIBLINGS", "ATTEND_SEMINAR")]
data[factor_columns] <- lapply(data[factor_columns], factor)
head(data)
```

## Exploring the Dataset

### Overview

Phew, that was some tedious cleaning part. Now time to quickly explore a little
about the data. 

```{r}
summary(data)
```


A rough looking at the summary shows that:

- There are only 145 rows, so there might not be enough data. 

- The students mostly prepare for their exam nearly last minute,

- and they don't seem to always listens to class. 

- Another surprising fact was that about half(64 / 145) of the students don't think
that the class is useful for their studies.

- Most of the students (74 + 29/ 145) spends lesser than 5 hours on studying weekly

## Splitting train / test set

I am going 25:75 for test:train data separation.

```{r}
set.seed(1)
train = sample(1:nrow(data), nrow(data) / 4 * 3)
```

## Using Random Forest

Random Forest is one of my favorite model, so I will quickly use a random forest 
model to model the dataset. Then, the Gini Index decrement will be used to assess how
important

```{r}
model.randomForest <- randomForest(GRADE ~ ., data = data, subset = train, mty = 13,
                                  importance = TRUE)



imp <- model.randomForest$importance[, "MeanDecreaseGini"]
imp <- as.data.frame(imp)
imp$varnames <- rownames(imp) 

ggplot(imp, aes(x=reorder(varnames, imp), weight=imp, fill=as.factor(imp))) + 
  geom_bar() +
  ylab("Mean Decrease in Gini Index") +
  xlab("Variable Name") +
  coord_flip() +
  dark_theme_gray() +
  theme(legend.position="none") +
  labs(title = "Mean Decrease in Gini Index for each variable")
```

Looking at the chart above, 
- It is not surprising that course ID will affect the 
grade the most because some courses are definitely easier and others harder,
so clearly everyone from the same course will have similar grades.

- Besides, cuml_gpa's and exp_gpa's effects are obvious too, as your past gpa is
higher, your grade will be higher because if you performed well in the past,
you will probably perform equally well in the future too.

- The impacts of MOTHER_EDU and FATHER_EDU seem to be significant as well. This is
naturally true as I am expecting the higher the parents' educations, the higher the
students' grades.

- The unexpected effect was from number of siblings which have
more impact than study hours. I guess when you have more
siblings means that you have to spend more time with siblings will result in 
worse grades? Time to verify this theory with a plot.

```{r}
data %>% 
  group_by(X._SIBLINGS, GRADE) %>%
  summarize(n = n()) %>%
  mutate(percentage = n / sum(n)) %>%
  ggplot(data, mapping = aes(x = X._SIBLINGS, fill = GRADE, y = percentage)) +
  geom_bar(position = "dodge", stat = "identity") + 
  dark_theme_gray() +
  labs(title = "Grade against Number of Siblings ") + 
  scale_y_continuous(labels = function(x) paste0(x*100, "%"))
```

From this graph, I can't exactly tell the relationship between Grade and X._Siblings.
There does seem to have a particularly high number of DC and DD when the student has
2 siblings somehow. On the other hand, when you have 5 or more siblings, it seems
unlikely to to get a grade AA due to distraction.

Meanwhile, relative to other numbers, having 4 siblings will make you more likely 
to get the best grade somehow. I don't really get this relationship here.

### Testing the model

```{r}
pred.randomForest <- predict(model.randomForest, data[-train, ])
table(pred.randomForest, data$GRADE[-train])
```

Hmmm, random forest's prediction doesn't seem to be particularly good here. It
only yields an accuracy of
`r sum(pred.randomForest == data$GRADE[-train]) / nrow(data[-train, ])`.
Slightly disappointed, let's try with another model.

## Using Support Vector Machine

### Linear Kernel

```{r}
set.seed(1)
tune.svm.linear <- tune(svm, GRADE ~ . , data = data[train, ], kernel = "linear",
                        ranges = list(cost = c(0.1, 1, 10, 100, 1000), gamma = c(0.5, 1, 2, 3, 4)))
model.svm.linear <- svm(GRADE ~ . , data = data, kernel = "linear", cost = 
                          tune.svm.linear$best.parameters$cost,
                        gamma = tune.svm.linear$best.parameters$gamma, subset = train)
pred.svm.linear <- predict(model.svm.linear, data[-train, ])
table(pred.svm.linear, data$GRADE[-train])
```

The accuracy of linear kernel for svm is 
`r sum(pred.svm.linear == data$GRADE[-train]) / nrow(data[-train, ])`.

### Polynomial Kernel

```{r}
tune.svm.poly <- tune(svm, GRADE ~ . , data = data[train, ], kernel = "polynomial",
                        ranges = list(cost = c(0.1, 1, 10, 100, 1000), gamma = c(0.5, 1, 2, 3, 4)))
model.svm.poly <- svm(GRADE ~ . , data = data, kernel = "polynomial", cost = 
                          tune.svm.poly$best.parameters$cost,
                        gamma = tune.svm.poly$best.parameters$gamma, subset = train)
pred.svm.poly<- predict(model.svm.poly, data[-train, ])
table(pred.svm.poly, data$GRADE[-train])
```

The accuracy of polynomial kernel for svm is 
`r sum(pred.svm.poly == data$GRADE[-train]) / nrow(data[-train, ])`.

### Radial Kernel

Finally, let's try with Radial Kernel


```{r}
tune.svm.radial <- tune(svm, GRADE ~ . , data = data[train, ], kernel = "radial",
                        ranges = list(cost = c(0.1, 1, 10, 100, 1000), gamma = c(0.5, 1, 2, 3, 4)))
model.svm.radial <- svm(GRADE ~ . , data = data, 
                        kernel = "radial",cost = 
                          tune.svm.radial$best.parameters$cost,
                        gamma = tune.svm.radial$best.parameters$gamma, subset = train)
pred.svm.radial <- predict(model.svm.radial, data[-train, ])
table(pred.svm.radial, data$GRADE[-train])
```

The accuracy of radial kernel for svm is 
`r sum(pred.svm.radial == data$GRADE[-train]) / nrow(data[-train, ])`.

Overall, the performance of the svm in this dataset seems to be quite poor, at least
from the perspective of test set accuracy. The highest accuracy of all SVM models
which is still lower than the accuracy from random forest model.

## Multinomial Logistic Regression

At this point I might as well test the logistic regression, although I am
not particularly hopeful about it.

```{r}
model.log <- multinom(GRADE ~ . , data = data, subset = train)
pred.log <- predict(model.log, newdata = data[-train, ])
table(pred.log, data$GRADE[-train])
```

The accuracy of multinomial logistic regression is 
`r sum(pred.log == data$GRADE[-train]) / nrow(data[-train, ])`.

Rather low as well, and this is totally expected since other more complex models
couldn't solve it. However, let's go for another similar 

## The Lasso

Also will try out the Lasso, as not every column might affect the GRADE.

```{r}
x <- data[train, ] %>%
  select(-GRADE)
y <- data$GRADE[train]
data.tmp <- cbind(x, y)
grid <- 10^seq(10,-2, length =100)
model.tmp <- model.matrix(y ~ ., data = data.tmp)

model.lasso <- cv.glmnet(model.tmp, y, alpha = 1, family = "multinomial", lambda = grid)

model.tmp <- model.matrix(y ~ ., data = data.tmp[-train, ])

x <- data[-train, ] %>%
  select(-GRADE)
y <- data$GRADE[-train]
data.tmp <- cbind(x, y)
model.tmp <- model.matrix(y ~ ., data = data.tmp)

pred.lasso <- predict(model.lasso, newx = model.tmp, s = model.lasso$lambda.min,
                      type = "class")

table(pred.lasso, y)
```

The accuracy of multinomial lasso regression is 
`r sum(pred.lasso == data$GRADE[-train]) / nrow(data[-train, ])`.

Not that bad. It was almost as good as random forest. Finally, let's go with the KNN.

## K Nearest Neighbor

```{r}
x <- data %>%
  select(-GRADE)
y <- data$GRADE

train.control <- trainControl(method = "repeatedcv", number = 10)

model.knn <- train(GRADE ~., data = data[train, ], method = "knn",
               trControl = train.control,
                preProcess = c("center", "scale"),
                tuneLength = 10,
               tuneGrid   = expand.grid(k = 1:10))

pred.knn <- predict(model.knn, x[-train, ])
table(pred.knn, y[-train])
```

```{r}
summary(model.knn)
```


The accuracy of knn is 
`r sum(pred.knn == data$GRADE[-train]) / nrow(data[-train, ])`.

### Note on KNN

When I was carrying out KNN using the training above, I came into this warning
(not just 1 or 2, but many of them)

```{r eval = FALSE}
These variables have zero variances: LIVINGother, MOTHER_JOBself-employment
```

It totally struck me dumb because I have checked that there are multiple cases
of exceptions regarding these columns.

After undertaking some research on my own ([source](https://rstatisticsblog.com/data-science-in-action/data-preprocessing/how-to-identify-variables-with-zero-variance/)), I found out that if the column contains too little of a class
will result in unstable prediction model. Thus, I guess this explains why
KNN's best k is 1.

Regardless, KNN did perform relatively well.


## Conclusion

Despite many different types of classification algorithms are used, some of them aren't even 
performing better than 30%, and the best ones are still below 40%.

In comparison to random guessing of 
`r 1 / length(unique(data$GRADE)) * 100`,
the model is of course much better, at an accuracy of approximately 2x random guessing.

I would say that a big factor was due to the size of dataset which after
split into train test, there's just slightly above 100 for the training set.
In this scenario, I can't do much as well to overcome this issue.

On the other hand, there could be much more reasons that explain the student's grade
like their lifestyle and other factors which may not be able to be measured easily.

Besides, I avoided PCA for dimentionality reduction because most of the columns
are just factors instead of numerical, so I decide to just avoid them.

Finally, here's a table of the prediction results

```{r echo=FALSE}
results <- data.frame(
  model = c("Random Forest", "Support Vector Machine - Linear",
           "Support Vector Machine - Polynomial", "Support Vector Machine - Radial",
           "Multinomial Logistic Regression", "The Lasso", "KNN"),
  result = c(sum(pred.randomForest == data$GRADE[-train]) / nrow(data[-train, ]),
             sum(pred.svm.linear == data$GRADE[-train]) / nrow(data[-train, ]),
             sum(pred.svm.poly == data$GRADE[-train]) / nrow(data[-train, ]),
             sum(pred.svm.radial == data$GRADE[-train]) / nrow(data[-train, ]),
             sum(pred.log == data$GRADE[-train]) / nrow(data[-train, ]),
             sum(pred.lasso == data$GRADE[-train]) / nrow(data[-train, ]),
             sum(pred.knn == data$GRADE[-train]) / nrow(data[-train, ]))
)

results
```

Out of all, random forest and KNN performed equally well at an 
accuracy of 35%. I didn't do any further optimization except for using
some cross validations to optimize the parameters.

Nevertheless, I am rather satisfied with this 35% accuracy for it is a 
multiclass classification.

























